---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CanonicalForms

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/colin-fraser/CanonicalForms/branch/master/graph/badge.svg)](https://app.codecov.io/gh/colin-fraser/CanonicalForms?branch=master)
[![R-CMD-check](https://github.com/colin-fraser/CanonicalForms/workflows/R-CMD-check/badge.svg)](https://github.com/colin-fraser/CanonicalForms/actions)
<!-- badges: end -->

`CanonicalForms` is an R package for ensuring that data sets conform to an expected format.

## Installation

You can install the development version of CanonicalForms from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("colin-fraser/CanonicalForms")
```

## Basic Example

A `CanonicalForm` allows you to check whether a dataset conforms to an expected format. The following code creates a simple CanonicalForm object for the dataset `cars`, as well as a pair of copies of the dataset, one that conforms to the expected format and one that does not.

```{r example, message=TRUE, warning=TRUE}
library(CanonicalForms)

cf <- canonical_form(
  object_class = "data.frame",
  col_names = c("speed", "dist"),
  col_classes = c("numeric", "numeric")
)

passing <- cars
failing <- cars |> 
  setNames(c("speed", "distance"))

passing |> 
  is_canonical(cf)  # checking whether `passing` corresponds to the form specified in cf

failing |> 
  is_canonical(cf)
```

## Extracting and storing a `CanonicalForm`

It can be a little bit tedious to type out the full canonical schema in the way shown above, especially for datasets with a large number of columns. For this reason, there is an `extract_canonical_form` function which will use a dataset as a template to create a `CanonicalForm` object, as well as a `to_r_code` method that writes the boilerplate code for initializing a new CanonicalForm.

```{r}
# the starwars dataset is a tibble with 14 columns
starwars <- dplyr::starwars
head(starwars)

# this uses the starwars dataset as a template to extract a CanonicalForm
cf <- extract_canonical_form(starwars)
to_r_code(cf) # and this writes the boilerplate R code to construct that form
```
## Checking if a dataset is Canonical during transformations

Suppose I have a pipeline that does the following transformations to the `starwars` dataset.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
starwars_small <- starwars |> 
  transmute(name, height, mass = as.integer(mass)) |> 
  rename_with(toupper)

swcf <- extract_canonical_form(starwars_small)
```

Now I have another script where I'm trying performing the same transformations. I can add a call to `check_canonical` at the end of the transformations to make sure that the pipeline does what I expect. `check_canonical` returns its input, but will raise a warning if the checks fail.

```{r}
starwars_small_2 <- starwars |> 
  select(name, height, mass) |> 
  check_canonical(swcf)
```

You can also set it to raise an exception rather than a warning.

```{r eval=FALSE, include=FALSE}
starwars_small_2 <- starwars |> 
  select(NAME = name, HEIGHT = height, MASS = mass) |> 
  check_canonical(swcf, behavior = 'stop')
```

If the pipeline returns the expected format, nothing visible will happen.

```{r}
starwars_small_2 <- starwars |> 
  transmute(NAME = name, HEIGHT = height, MASS = as.integer(mass)) |> 
  check_canonical(swcf, behavior = 'stop')
head(starwars_small_2)
```

## Adding more checks

By default, a newly created `CanonicalForm` objects have three checks: they'll check that the type of dataset matches, the column names match, and the column types match. The package also provides other checks that can be run with `is_canonical` and `check_canonical`, and it's easy to write custom checks and add those as well.

```{r message=TRUE, warning=TRUE}
# passing example
swcf2 <- swcf |> 
  add_checks(
    # check that no NAME values are NA:
    no_nas = check_no_nas(cols = c('NAME')),  
    # check that HEIGHT and MASS are greater than 0:
    positive_values = check_greater_than(HEIGHT = 0,MASS = 0)
    )

starwars_small |> 
  check_canonical(swcf2)

# failing example
swcf3 <- swcf |> 
  add_checks(no_nas = check_no_nas(c("NAME", "HEIGHT", "MASS")),
             min_values = check_greater_than(HEIGHT = 1000, MASS = 0))

starwars_small |> 
  check_canonical(swcf3)
```

